A new metric for automatic discovery of periodic patterns in time series

### 2021-03-02 paper/main.Rmd
introduction done except timeseer

\noindent Take the example of the calendar display of electricity smart meter data in Figure (\ref{fig:calendar-elec}) used in @wang2020calendar for four households in Melbourne, Australia. The authors show how hour-of-the-day interact with weekdays and weekends and then move on to use calendar display to show daily schedules. The calendar display has several components in it, which helps us to look at energy consumption across hour-of-the-day, day-of-the-week, week-of-the-month, and month-of-the-year at once. Some interaction of these cyclic granularities, for example, how  day-of-week relates to month-of-year, could also be interpreted from this display. This is a great way for having an overview of energy consumption. However, if one wants to understand the periodicities in energy behavior and how the periodicities interact in greater detail, it is not easy to comprehend the interactions of some periodicities' from this display, due to the combination of linear and cyclic representation of time. For example, this display might not be the best to understand how hour-of-the-day or month-of-year varies across week-of-the-month as well as with each other. Furthermore, it is not clear what all interactions of cyclic granularities should be read from this display as there could be many combinations that one can look at. Moreover, _"calendar effects are not restricted to conventional day-of-week or month-of-year deconstructions"_ (@Gupta2020-vo) and could include other cyclic granularities like hour-of-week or day-of-fortnight, which could potentially become useful depending on the context.

@@

Additionally, even when it is known which all interactions to look at, all of them would not be interesting and that too will vary across different households. For example, area distribution quantiles are plotted for household 2 and 4 in Figure \ref{fig:id2}a and b respectively across month-of-year on facets and hour-of-day on x-axis. For the first household, the 75th and 90th percentile for January and February are very close, implying that energy usage for these months are generally on a much higher side, possibly due to the usage of air conditioners (January and February are peak summer in Australia), however for other months (Autumn and winter), the difference between the percentiles are not high, implying this household do not use so much heater as compared to air conditioner. Also, a lot of households in Victoria use gas heating and hence the usage of heaters might not be reflected here. The energy consumption for household 2 is also higher in summer months relative to autumn or winter, but the 75th and 90th percentile are far apart in all months, implying that the second household resorts to air conditioners much less regularly than the first one. Moreover, the 75th percentile distribution does not follow the same pattern as 90th percentile for all months for the first household, whereas, the pattern looks pretty similar for all months for the second household. Difference in the energy consumption could vary both across month-of-year (facets) and hour-of-day (x-axis). For the first household, both these cyclic granularities would deem important. Although, it seems like energy consumption across hours of the day are not that different across different months for the second household or atleast differences seem to be more prominent across month-of-year (facets) than hour-of-day (x-axis). It could be immensely useful to make the transition from all possible ways to only ways that could potentially be informative given a household.

@@

<!-- what is the dimension of the problem -->
The paper @Gupta2020-vo describes how we can compute all possible combinations of cyclic time granularities. If we have $n$ periodic linear granularities in the hierarchy table, then $n(n-1)/2$ circular or quasi-circular cyclic granularities could be constructed. Let $N_C$ be the total number of contextual circular, quasi-circular and aperiodic cyclic granularities that can originate from the underlying periodic and aperiodic linear granularities. The mapping of the graphical elements chosen in the paper implies that, for a numeric response variable, the graphics display distributions across combinations of cyclic granularities, one placed at x-axis and the other on the facet. That essentially implies there are $^{N_C}P_2$ possible pairwise plots exhaustively, where each plot would display a pair of cyclic granularities. This is large and overwhelming for human consumption.

<!-- Scagnostics literature -->
This is similar to Scagnostics (Scatterplot Diagnostics) by @tukey1988computer, which is used to discern meaningful patterns in large collections of scatterplots. Given a set of $v$ variables, there are $v(v-1)/2$ pairs of variables, and thus the same number of possible pairwise scatterplots. Therefore
for even small $v$, the number of scatterplots can be
large, and scatterplot matrices (SPLOMs) could easily run out of pixels when presenting high-dimensional data. @Dang2014-tw and @wilkinson2005graph provides potential solutions to this, where few characterizations help us to locate anomalies for defining several measures aimed to detect anomalies in density, shape, trend, and
other features in the 2D point scatters.

<!-- harmonies and why it is not enough -->
The paper (@Gupta2020-vo) narrows down the search from $^{N_C}P_2$ plots by identifying pairs of granularities that can be meaningfully examined together (a "harmony"), or when they cannot (a "clash"). However, even after excluding clashes, the list of harmonies left could be enormous for exhaustive exploration. Hence, there is a need to reduce the search even further by including only those harmonies which are informative enough. Also, ranking the remaining harmony pairs based on how well they capture the variation in the measured variable could be potentially useful.

In this paper, we aim to build a new measure to follow through these two main objectives:

- To choose harmonies for which distributions of categories are significantly different 
- To rank the selected harmonies from highest to lowest variation in the distribution of their categories.

@@

# A distance measure for quantifying patterns in harmonies

@@

# Normalization for the number of comparisons

@@

# Ranking and selecting significant harmonies

@@

## Results

@@

# Simulation environment

@@

# Application to residential smart meter dataset {#sec:application}

@@

A context when it could be useful is monitoring heart rates. There are devices that can detect each heartbeat and transfer the data to a receiver such as a watch or phone. These data could be available for a temporal scale as fine as a minute and it could be of interest to see regular patterns across any deconstruction of time coarser than a minute.

>>>>

\noindent Take the example of the calendar display of electricity smart meter data in Figure (\ref{fig:calendar-elec}) used in @wang2020calendar for four households in Melbourne, Australia. The authors show how hour-of-the-day interact with weekdays and weekends and then move on to use calendar display to show daily schedules. The calendar display has several components in it, which helps us to look at energy consumption across hour-of-the-day, day-of-the-week, week-of-the-month, and month-of-the-year at once. Some interaction of these cyclic granularities, for example, how  day-of-week relates to month-of-year, could also be interpreted from this display. This is a great way for having an overview of energy consumption. However, if one wants to understand the periodicities in energy behavior and how the periodicities interact in greater detail, it is not easy to comprehend the interactions of some periodicities' from this display, due to the combination of linear and cyclic representation of time. For example, this display might not be the best to understand how hour-of-the-day or month-of-year varies across week-of-the-month as well as with each other. Furthermore, it is not clear what all interactions of cyclic granularities should be read from this display as there could be many combinations that one can look at. Moreover, _"calendar effects are not restricted to conventional day-of-week or month-of-year deconstructions"_ (@Gupta2020-vo) and could include other cyclic granularities like hour-of-week or day-of-fortnight, which could potentially become useful depending on the context. Possible areas where it is useful are monitoring heart rates which could record number of heartbeats every minute or analyze web search data for which data is available for a temporal scale as fine as second.

<!-- context when it could be useful is monitoring heart rates. There are devices that can detect each heartbeat and transfer the data to a receiver such as a watch or phone. These data could be available for a temporal scale as fine as a minute and it could be of interest to see regular patterns across any deconstruction of time coarser than a minute. -->

@@

Even when it is known which all interactions to look at, all of them would not be interesting and that too will vary across different households. For example, area distribution quantiles are plotted for household 2 and 4 in Figure \ref{fig:id2}a and \ref{fig:id2}b respectively across month-of-year on facets and hour-of-day on x-axis. For the first household, the 75th and 90th percentile for January and February are very close, implying that energy usage for these months are generally on a much higher side, possibly due to the usage of air conditioners (January and February are peak summer in Australia), however for other months (Autumn and winter), the difference between the percentiles are not high, implying this household do not use so much heater as compared to air conditioner. Also, a lot of households in Victoria use gas heating and hence the usage of heaters might not be reflected here. The energy consumption for household 2 is also higher in summer months relative to autumn or winter, but the 75th and 90th percentile are far apart in all months, implying that the second household resorts to air conditioners much less regularly than the first one. Moreover, the 75th percentile distribution does not follow the same pattern as 90th percentile for all months for the first household, whereas, the pattern looks pretty similar for all months for the second household. Difference in the energy consumption could vary both across month-of-year (facets) and hour-of-day (x-axis). For the first household, both these cyclic granularities would deem important. Although, it seems like energy consumption across hours of the day are not that different across different months for the second household or atleast differences seem to be more prominent across month-of-year (facets) than hour-of-day (x-axis). It could be immensely useful to make the transition from all possible ways to only ways that could potentially be informative.

@@

<!-- If we have $n$ periodic linear granularities in the hierarchy table, then $n(n-1)/2$ circular or quasi-circular cyclic granularities could be constructed.  -->

<!-- what is the dimension of the problem -->
The paper @Gupta2020-vo describes how we can compute all possible combinations of cyclic time granularities. Let $N_C$ be the total number of contextual circular, quasi-circular and aperiodic cyclic granularities that can originate from the underlying linear granularities. The graphical mapping is such that distributions of a numeric response variable is displayed across combinations of cyclic granularities, one placed at x-axis and the other on the facet. That essentially implies there are $^{N_C}P_2$ possible pairwise plots exhaustively, where each plot would display a pair of cyclic granularities. This is large and overwhelming for human consumption.<!-- Scagnostics literature --> This problem is similar to Scagnostics (Scatterplot Diagnostics) by @tukey1988computer, which is used to discern meaningful patterns in large collections of scatterplots. Given a set of $v$ variables, there are $v(v-1)/2$ pairs of variables, and thus the same number of possible pairwise scatterplots. Therefore, even for small $v$, the number of scatterplots can be large, and scatterplot matrices (SPLOMs) could easily run out of pixels when presenting high-dimensional data. @Dang2014-tw and @wilkinson2005graph provides potential solutions to this, where few characterizations help us to locate anomalies in density, shape, trend, and other features in the 2D point scatters.

<!-- harmonies and why it is not enough -->
This work is a natural extension of our work (@Gupta2020-vo), which narrows down the search from $^{N_C}P_2$ plots by identifying pairs of granularities that can be meaningfully examined together (a "harmony"), or when they cannot (a "clash"). However, even after excluding clashes, the list of harmonies left could be enormous for exhaustive exploration. Hence, there is a need to reduce the search even further by including only those harmonies which are informative enough. @inference and @Majumder2013-hb present some methods to quantify the strength of pattern and noise through visual inference, similar to numerical testing. But this is an evolving field as human cognition (which act as the statistical tests in visual inference) might vary across humans even for the same plots. In this paper, we build a new distance measure which could be used to detect automatically detect significant periodic patterns.

Our contributions in this paper are:

 * We introduce a new distance measure for detecting periodic interactions. This induces data reduction which allows for identification of patterns, if any, in the time series data.

 * We show that the distance metric could be used to rank the periodic patterns based on how well they capture the variation in the measured variable as they have been normalized for different number of comparisons.

 * We device a framework for choosing a threshold, which will result in detection of only significantly interesting periodic patterns in the time series data.

The article is organized as follows. Section \ref{sec:computation-wpd} introduces a new distance measure, discusses the reasoning behind choosing such a measure and presents some results to study the behavior of the measure. Section \ref{sec:norm-wpd} describes a methodology to normalize the distance measure so that it can qualify as a measure that can be compared across different comparisons and datasets. Section \ref{sec:rank-wpd} discusses how to choose a threshold to select only significant harmonies. Section \ref{sec:application-wpd} presents an application to a residential smart meter data in Melbourne to show how this distance measure acts as a way to automatically detect periodic patterns in time series.

<!-- Also, ranking the remaining harmony pairs based on how well they capture the variation in the measured variable could be potentially useful.   -->

@@

# A distance measure for quantifying patterns in harmonies {#sec:computation-wpd}

@@

# Normalization for the number of comparisons {#sec:norm-wpd}

@@

# Ranking and selecting significant harmonies {#sec:rank-wpd}

@@

## Simulation design

@@

## Simulation environment

@@

# Application to residential smart meter dataset {#sec:application-wpd}

@@

 A common remedy would be to zoom into “interesting” segments, but how do we identify “interesting” segments?
